### 2022-06-20 draft.tex
Deploying to gh-pages from @ PoisotLab/ms_betacov_mapping@bdd862abac2f96426c34cd878e05be6d7808831f 🚀

Spillover risk is complex and, within a pool of susceptible wildlife
hosts, is driven by a multiplicity of factors (Plowright et al. 2017).
Although proxies for the local richness of hosts are commonly analysed
(Anthony et al. 2017, Ruiz-Aravena et al. 2022), these approaches
deliberately oversimplify interspecific heterogeneity in immunity,
behavior, and contact with humans. At the coarsest level, some species
play a more central role in the host-virus network because of key
functional traits (for which phylogeny is a reasonable proxy) (Albery et
al. 2020); these include, notably, body mass, life history, diet, and
use for human-occupied spaces Olival et al. (2017). Global maps of
spillover risk often struggle to distill these features into
interpretable risk maps, and overlook highly-unique host pools that
allow for the rapid evolution of highly divergent viruses (Agosta et al.
2010). In the case of generalist pathogens like betacoronaviruses, there
is conceptual and empirical support to the idea that these
community-level mechanisms are even more important in driving the
overall risk (Power and Mitchell 2004). In this paper, we examine the
biogeographic structure of bat-betacoronavirus associations by adopting
an analysis framework grounded in concepts from the Geographic Mosaic
Theory of Coevolution (GMTC; Thompson 2005). The GMTC posits that
coevolutionary processes between pairs (Thompson 1994) or complexes
(Janzen 1980) of species are structured in space by the rippling effects
of abiotic conditions onto evolutionary mechanism, resulting in
spatially fragmented evolutionary dynamics, coupled only by
dispersal-related processes (Gomulkiewicz et al. 2000). In turn, these
spatially fragmented processes can lead taxonomically homogeneous
systems to have different structure and dynamics over large spatial
extents (see \emph{e.g.} Price 2002).

The GMTC predicts a spatial fragmentation of coevolutionary dynamics
under the joint action of three processes (see notably Gomulkiewicz et
al. 2007), which all have the potential to act on outbreak potential,
pathogen tramission, and disease virulence (Parratt et al. 2016, Turner
et al. 2021). First hot- and coldspots of coevolution can appear when
the intensity of \emph{interaction} (in terms of reciprocal fitness
consequences) varies spatially, because of \emph{e.g.} partial range
overlap between organisms (Nuismer et al. 2003). Hot- and coldspots are
notoriously difficult to identify in nature (Laine 2005), especially
when the structure of range overlap is complex, and the reciprocal
fitness effects are relatively low---both of which conditions are met in
the bat-betacoronavirus complex. Second, the GMTC supposes the existence
of selection mosaics, wherein the intensity of \emph{selection} varies
across space; the strength of reciprocal selection responds to the
biotic complexity of the community (locally diverse hosts and viruses
are more biotically complex; Thrall et al. 2007) \emph{and} to the local
favorability of the environment (Hochberg and Baalen 1998). In a system
with near global distribution and variations in relative richness, like
the bat-betacoronavirus complex, the GMTC predicts the establishment of
many different coevolutionary trajectories (and resulting viral
regimes). The final GMCT process is trait remixing, under which changes
in coevolutionary dynamics happen due to the arrival (or departure) of
functional traits, through changes in community composition due to
invasions, meta-community dynamics, and disperal; these changes are
likely to happen in overlap zones between ``long branches'' of
coevolution.

We turn the processes on the GMTC into definitions of spillover risk
from viruses to hosts (focusing on the bats-betacoronavirus complex),
with a specific emphasis on the potential to create independant
coevolutionary dynamics (and therefore spatial fragmentation in the
risk) through heterogeneity. Our components of risk are (i) the
phylogenetic diversity of hosts, representing the diversification
potential of the system (hotspots, selection mosaics), under the
assumption that bat diversification is not primarily driven by viruses
(Gorbunova et al. 2020); (ii) viral sharing among hosts, representing
levels of viral exchange within host communities that may act as a
buffer against viral specialization in any given host (i.e., high
sharing creates coldspots of selective pressures from any given host
species); and (iii) the local uniqueness of the bat community,
representing the potential for viruses to be exposed to novel host
traits in unusual combinations (trait remixing). We turn these features
into a spatially-explicit additive mapping of zoonotic risk components,
revealing the extreme heterogeneity of risk at the global scale.
Explicitely framing the notion of spillover risk based on propositions
from the GMTC (which is to say, based on a framework linking
interactions between species to change within species) is a novel idea,
that should be relatively general. Indeed, it only assumes the action of
well described evolutionary mechanisms. The benefit of this approach is
to provide the potential for a more dynamic and nuanced understanding of
risk: not only on ecological timescales, but also by providing clues
about which areas can change over micro-evolutionary timescales. This
provides a way to look at spatial structure by accounting for more
notions than species richness/similarity, but also a way to identify
spatial areas of higher risk.

@@

@@

\leavevmode\hypertarget{ref-Albery2022UrbMam}{}%
Albery, G. F. et al. 2022. Urban-adapted mammal species have more known
pathogens. - Nature Ecology \& Evolution 6: 794--801.

@@

\leavevmode\hypertarget{ref-Laine2005SpaSca}{}%
Laine, A.-L. 2005. Spatial scale of local adaptation in a plant-pathogen
metapopulation. - Journal of Evolutionary Biology 18: 930--938.

>>>>

Spillover risk is complex and, even within a pool of susceptible
wildlife hosts, driven by a multiplicity of factors (Plowright et al.
2017). Although host richness is often used as a coarse proxy for
hotspots of emergence risk (Anthony et al. 2017, Ruiz-Aravena et al.
2022), these approaches deliberately oversimplify interspecific
heterogeneity in immunology, behavior, and other traits. Global maps of
spillover risk often struggle to distill these features into
interpretable risk maps, and overlook highly-unique host pools that
allow for the rapid evolution of highly divergent viruses (Agosta et al.
2010). In the case of generalist pathogens like betacoronaviruses, there
is conceptual and empirical support to the idea that these
community-level mechanisms are even more important in driving hotspots
of overall risk (Power and Mitchell 2004).

These kinds of dynamics lend themselves to interpretation through the
lens of the Geographic Mosaic Theory of Coevolution (GMTC; Thompson
2005), which connects microevolutionary dynamics in symbiotic
interactions to macroecological dynamics in host communities. The GMTC
posits that coevolutionary processes between pairs (Thompson 1994) or
complexes (Janzen 1980) of species are structured in space by the
rippling effects of abiotic conditions onto evolutionary mechanism,
resulting in spatially fragmented evolutionary dynamics, coupled only by
dispersal-related processes (Gomulkiewicz et al. 2000). In turn, these
spatially fragmented processes can lead taxonomically homogeneous
systems to have different structure and dynamics over large spatial
extents (see \emph{e.g.} Price 2002). The GMTC predicts a spatial
fragmentation of coevolutionary dynamics under the joint action of three
processes (see notably Gomulkiewicz et al. 2007), which all have the
potential to act on outbreak potential, pathogen tramission, and disease
virulence (Parratt et al. 2016, Turner et al. 2021). First hot- and
coldspots of coevolution can appear when the intensity of
\emph{interaction} (in terms of reciprocal fitness consequences) varies
spatially, because of \emph{e.g.} partial range overlap between
organisms (Nuismer et al. 2003). Second, the GMTC supposes the existence
of selection mosaics, wherein the intensity of \emph{selection} varies
across space; the strength of reciprocal selection responds to the
biotic complexity of the community (locally diverse hosts and viruses
are more biotically complex; Thrall et al. 2007) \emph{and} to the local
favorability of the environment (Hochberg and Baalen 1998). Finally,
trait remixing occurs when coevolutionary dynamics are driven by the
arrival (or departure) of functional traits, through changes in
community composition due to invasions, meta-community dynamics, and
disperal.

Each of these elements can be applied to a relatively quantifiable
aspect of host-virus ecology: (i) viral sharing among hosts,
representing the strength of potential interaction between viruses and
any one host (i.e., places where viruses undergo constant host switching
may be coevolutionary coldspots); (ii) the phylogenetic diversity of
hosts, as a proxy for variation in the immunological mechanisms that
antagonize viruses (i.e., the selection mosaic); and (iii) the local
uniqueness of the bat community, representing the potential for viruses
to be exposed to novel host traits (e.g., variation in receptor
sequences). Together, we argue that these can be used to identify the
evolutionary drivers that---in conjunction with transmission processes
(e.g., viral prevalence in reservoirs and animal-human contact rates)---
determine disease emergence risk.

Here,

We turn the processes on the GMTC into definitions of spillover risk
from viruses to hosts (focusing on the bats-betacoronavirus complex),
with a specific emphasis on the potential to create independent
coevolutionary dynamics (and therefore spatial fragmentation in the
risk) through heterogeneity.

@@

RELOCATE: Indeed, it only assumes the action of well described
evolutionary mechanisms. The benefit of this approach is to provide the
potential for a more dynamic and nuanced understanding of risk: not only
on ecological timescales, but also by providing clues about which areas
can change over micro-evolutionary timescales.

@@

@@
